---
title: "Geoeikonal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Geoeikonal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 7,
  fig.height = 4,
  comment = "#>"
)
```

The `geoeikonal` package computes generalized distances and arrival times on a geographic grid by
solving the eikonal equation
$$
| \nabla D | = C
$$

by the fast marching algorithm of Sethian (1996). If $C$ represents the local cost of travel, then 
$D$ represents the cost weighted distance from a set of source points in the grid.  Alternately if 
$C=S^{-1}$ where $S$ is the local speed of travel, then $D$ can be interpreted as an arrival time.

```{r setup}
library(geoeikonal)
```


## Example

The `rsteikonal` function is analogous to the `gridDist` and `costDist` functions from the terra
package.


Given a raster of travel costs with source points marked as zero, 
```{r}
library(terra)
cost <- rast(res=1,val=1)
src <- cbind(-90.5,-30)
values(cost)[cellFromXY(cost,src)] <- 0
```

for each point in the raster `rsteikonal` computes the minimal weighted distance to a source point 
assuming a spherical Earth.
```{r}
d.eikonal <- rsteikonal(cost)
plot(d.eikonal)
```

This is similar to `gridDist` from the terra package
```{r}
d.griddist <- gridDist(cost)
plot(d.griddist)
```

The two differ in that `gridDist` assumes an ellipsoidal Earth, and `gridDist` computes the distance
along paths that are constrained to pass through the centers of the intervening cells. This 
constraint induces facets in the computed distance surface
```{r}
plot(terrain(d.griddist,v="aspect"),col=hcl((1:360)%%360,180,65))
```

In contrast, `rsteikonal` computes distances along paths that do not necessarily pass through the 
centers of the grid cells, and the distance surface is smoother
```{r}
plot(terrain(d.eikonal,v="aspect"),col=hcl((1:360)%%360,180,65))
```

## Barriers

Cells with a missing, non-finite or negative cost are treated as impenetrable barriers. 

Create a raster map in which which the ocean has cost one but land is `NA`, with a source point in
the South Pacific
```{r}
library(rnaturalearth)
library(rnaturalearthdata)
cost <- rast(res=0.1,val=1)
src <- cellFromXY(cost,cbind(-90.5,-30))
values(cost)[src] <- 0
## Extract land polygons with `rnaturalearth`
land <- rasterize(vect(ne_countries(scale="medium",returnclass="sf")), cost, field = 1)
cost[!is.na(land)] <- NA  # Set land cells to NA
plot(cost)
```

In this case, `rsteikonal` computes the minimal distance to the source point along paths constrained 
to lie in the ocean
```{r}
d.eikonal <- rsteikonal(cost)
plot(d.eikonal)
```

The fast marching algorithm employed by `rsteikonal` relies on a four point neighbourhood, and so 
one cell can be reached from another only if they are connected by a sequence of cells for which 
each pair of successive cells share a common edge. In this example, the Black sea is not connected 
to the Mediterranean and is unreachable from the source.  In contrast, `gridDist` uses an eight 
point neighbourhood, and at this resolution the Black sea is connected to the Mediterranean and can 
be reached from the source.
```{r}
d.griddist <- gridDist(cost)
plot(d.griddist)
```


## gradient

The gradient of the distance surface can be computed with the `rstgradient`
```{r}
g.eikonal <- rstgradient(d.eikonal)
plot(g.eikonal)
```

This computes the same "upwind" gradient used in the fast marching algorithm, and so the modulus 
of the gradient reproduces the cost function for cells that are reachable from the sources
```{r}
plot(sqrt(g.eikonal[[1]]^2 + g.eikonal[[2]]^2))
```






