---
title: "Geoeikonal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Geoeikonal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 7,
  fig.height = 4,
  comment = "#>"
)
```

The `geoeikonal` package computes generalized distances and arrival times on a geographic grid by
solving the eikonal equation
$$
| \nabla D | = C
$$

by the fast marching algorithm of Sethian (1996). If $C$ represents the local cost of travel, then 
$D$ represents the cost weighted distance from a set of source points in the grid.  Alternately if 
$C=S^{-1}$ where $S$ is the local speed of travel, then $D$ can be interpreted as an arrival time.

```{r setup}
library(geoeikonal)
```


## Example

The `rsteikonal` function is analogous to the `gridDist` and `costDist` functions from the terra
package.


Given a raster of travel costs with source points marked as zero, 
```{r}
library(terra)
cost <- rast(res=1,val=1)
src <- cbind(-90.5,-30)
values(cost)[cellFromXY(cost,src)] <- 0
```

for each point in the raster `rsteikonal` computes the minimal weighted distance to a source point 
assuming a spherical Earth.
```{r}
d.eikonal <- rsteikonal(cost)
plot(d.eikonal)
```

This is similar to `gridDist` 
```{r}
d.griddist <- gridDist(cost)
plot(d.griddist)
```

`gridDist` assumes an ellipsoidal Earth and the minimal paths are constrained to pass through the 
centers of the intervening cells. As a consequence, the gradient of the distance surface computed by
`gridDist` is not smooth 
```{r}
plot(terrain(d.griddist,v="aspect"),col=hcl((1:360)%%360,180,65))
```

In contrast, the discontinuities in the gradient of the distance surface computed by `rsteikonal` 
are determined by the cost function 
```{r}
plot(terrain(d.eikonal,v="aspect"),col=hcl((1:360)%%360,180,65))
```

## Barriers

Cells with a cost that is missing, not finite or negative are treated as impenetrable barriers. 


Create a raster map in which which the ocean has cost one but land is `NA`.
```{r}
library(rnaturalearth)
library(rnaturalearthdata)
cost <- rast(res=0.1,val=1)
src <- cellFromXY(cost,cbind(-90.5,-30))
values(cost)[src] <- 0
## Extract land polygons with `rnaturalearth`
land <- rasterize(vect(ne_countries(scale="medium",returnclass="sf")), cost, field = 1)
cost[!is.na(land)] <- NA  # Set land cells to NA
plot(cost)
```

The `rsteikonal` computes the minimal distance to the source point along paths constrained to lie in
the ocean
```{r}
d.eikonal <- rsteikonal(cost)
plot(d.eikonal)
```

The fast marching algorithm employed by `rsteikonal` relies on a four point neighbourhood, and so 
a cell can be reached from another only if they are connected by a sequence of cells that share a 
common edge. So in this example, the Black sea is not connected to the Mediterranean and is 
unreachable from the source.  In contrast, `gridDist` uses an eight point neighbourhood and at this 
resolution the Black sea is connected to the Mediterranean and can be reached from the source.
```{r}
d.griddist <- gridDist(cost)
plot(d.griddist)
```
